The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-07-25 manuscript.Rmd
incorporate GP's edits

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists. We filtered the results to only include the same higher taxonomic level taxa as in the faunistic lists, and kept only records for which the species could be matched to a species name in WoRMS. For the Gulf of Mexico, we retrieved  `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species in iDigBio (from $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records), and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_gom_records), big.mark=",")` records). For the Pacific Northwest, we retrieved `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` species in iDigBio ($n=$ and `r format(nrow(idigbio_pnw_records), big.mark=",")`) and `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_pnw_records), big.mark=",")`). Overall, the number of species known from the Gulf of Mexico match the number estimated by experts (`r format(11150-1541-434, big.mark=",")` species [@Fautin2010]). For the Pacific coast, this number is lower than the estimate (`r format(8374-909-134, big.mark=",")` species [@Fautin2010]), but this estimate included species from California and from deeper areas that were not included in Kozloff's faunistic list.

For the Gulf of Mexico, the majority of the species included in the expert faunistic list are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-coverage)A). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)B). Actually, the majority of the species in the faunistic list are not recorded in any of the public databases. Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.

@@

We investigated the number of species for each phylum that contributed to the discrepancies between the regional species lists and the database records. We found that the better studied phyla contributed the most to the differences across the data sources. OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list (Fig. \@ref(fig:database-overlap)). Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have no digital records in these geographic areas.

@@

Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced data points and therefore were not returned in our queries.

Untangling the relative contribution of these factors to explain the discrepancies is difficult as it would require to know the criteria used by the experts who have assembled the lists. However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepancy, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico. In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.

@@

We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species from iDigBio. Most samples were collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A). For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline. For the West coast, the Pacific Northwest, the San Francisco Bay Area, and Southern California have the highest richness (Fig. \@ref(fig:map-diversity-sampling)B).

@@

These patterns of spatial distribution for the species richness are strongly confounded by the sampling effort. Investigating the relation between the number of samples and the number of species for each cell of the map (Fig. \@ref(fig:correlation-sampling)) reveals that it takes about 100 records per cell before there is clear redundancy in the sampled species (each record represent a different species).

@@

The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record. Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century led to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species. Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. While less explored than the US, a recent study in Western Australia reported that the collection of 135 locations led to the collection of 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014], highlighting the potential for discovery in the oceans.

>>>>

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists. We filtered the results to only include the same higher taxonomic level taxa as in the faunistic lists, and kept only records for which the species could be matched to a species name in WoRMS. For the Gulf of Mexico, we retrieved  `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species in iDigBio (from $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records), and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_gom_records), big.mark=",")` records). For the Pacific Northwest, we retrieved `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` species in iDigBio ($n=$ and `r format(nrow(idigbio_pnw_records), big.mark=",")` records) and `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records). Overall, the number of species known from the Gulf of Mexico match the number estimated by experts (`r format(11150-1541-434, big.mark=",")` species [@Fautin2010]). For the Pacific coast, this number is lower than the estimate (`r format(8374-909-134, big.mark=",")` species [@Fautin2010]), but this estimate included species from California and from deeper areas that were not included in Kozloff's faunistic list.

<!-- this paragraph needs overhaul to make it clearer -->
For the Gulf of Mexico, the majority of the species included in the expert faunistic list are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-coverage)A). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)B). Actually, the majority of the species in the faunistic list are not recorded in any of the public databases. Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.

@@

<!-- need to separate across all phyla statements and by phylum statements -->
We investigated the number of species for each phylum that contributed to the discrepancies between the regional species lists and the database records. We found that the more diverse phyla contributed the most to the differences across the data sources. OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list (Fig. \@ref(fig:database-overlap)). Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have no digital records in these geographic areas.

@@

<!-- need to separate better baises for which I have data and those for which I don't -->
Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonymized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced data points and therefore were not returned in our queries.

Untangling the relative contribution of these factors to explain the discrepancies is difficult as it would require to know the criteria used by the experts who have assembled the lists. However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio across all areas, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepancy, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico. In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.

@@

<!-- add sentence
We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species from iDigBio. Most samples were collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A). For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline. For the West coast, the Pacific Northwest, the San Francisco Bay Area, and Southern California have the highest richness (Fig. \@ref(fig:map-diversity-sampling)B).

@@

Spatial patterns of species richness recovered are strongly confounded by sampling effort. Samples of <100 records per cell commonly have as many species as records (Fig. \@ref(fig:correlation-sampling)).

@@

Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record in iDigBio. Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century led to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species. Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. While less explored than the US, a recent study in Western Australia reported that the collection of 135 locations led to the collection of 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014], highlighting the potential for discovery in the oceans.
